Range Analysis of Whole Programs

### 2012-03-21 doc/Manuscript/paper.tex
Added Victor's fixes to the paper.

% Our contribution.
In this paper we provide a complete description of a range analysis algorithm,
and show extensive experimental data that justifies our engineering choices.
Our first algorithmic contribution on top of previous works is a three-phases
approach to handle comparisons between variables without resorting to any
exponential time technique.
The few publicly available implementations of range analyses that we are
aware of, such as those in FLEX~\footnote{The MIT's FLEX/Harpoon compiler
provides an implementation of Stephenson's algorithm~\cite{Stephenson00}, and
is available at \texttt{http://flex.cscott.net/Harpoon/}.} and
gcc~\footnote{Gcc's VRP pass implements a variant of Patterson's
algorithm~\cite{Patterson95}.} only deal with comparisons between variables and
constants.
Even theoretical works, such as Su and Wagner's~\cite{Su05} or Gawlitza
{\em et al.}'s~\cite{Gawlitza09} suffer from this limitation.
This deficiency is one of the reasons explaining why none of these works has
made their way into industrial-strength compilers.
Two other insights allow our implementation to scale up to very large
programs.
We use Bodik's Extended Static Single Assignment (e-SSA) form~\cite{Bodik00} to
perform range analysis sparsely.
This program representation ensures that the interval associated with
a variable is constant along its entire live range, and allows us to compute
path-sensitive results.
Finally, we process the strongly connected components that underline our
constraint system in topological ordering.
It is well-know that this technique is essential to speedup constraint solving
algorithms~\cite[Sec 6.3]{Nielson99}; however, we show that a careful
propagation of information along strong components not only gives us speed, but
also improves the precision of our results.

@@

In this section we explain the algorithm that we use to solve the range
analysis problem.
This algorithm involves a number of steps.
First, we convert the program to a suitable intermediate representation that
makes it easier to extract constraint.
From these constraints, we build a dependence graph that allows us to do
range analysis sparsely.
Finally, we solve the constraints applying different fix-point iterators on
this dependence graph.
Figure~\ref{fig:algorithm} gives a global view of this algorithm.
Some of the steps in the algorithm are optional.
They improve the precision of the range analysis, at the expenses of a longer
running time.

@@

\paragraph{Choosing a Program Representation.}
The solution that we found to the range analysis problem in Figure~\ref{fig:ex1}
is imprecise because we did not take conditional tests into considerations.
Branches give us information about the ranges that some variables assume, but
only at {\em specific} program points.
For instance, given a test such as $(k_1 < 100)?$ in  Figure~\ref{fig:ex1}(b),
we know that $I[k_1] \sqsubseteq [-\infty, 99]$ whenever the condition is true.
In order to encode this information, we might split the live range of $k_1$
right after the branching point; thus, creating two new variables, one at the
path where the condition is true, and another where it is false.
There is a program representation, introduced by Bodik
{\em et al.}~\cite{Bodik00}, that performs this live range splitting:
the {\em Extended Static Single Assignment} form, or e-SSA for short.

We use a non-iterative algorithm to convert a program to e-SSA form.
Let $(v < c)?$ be a conditional test, and let $l_t$ and $l_f$ be labels in
the program, such that $l_t$ is the target of the test if the condition is true,
and $l_f$ is the target when the condition is false.
If $l_f$ dominates any use of $v$, then we insert at $l_f$ a copy
$v_f = v \sqcap [-\infty, c-1]$, where $v_f$ is a fresh name.
We then rename every use of $v$ that is dominated by $l_f$ to $v_f$.
Dually, if $l_t$ dominates any use of $v$, then we insert at $l_t$ a copy
$v_t = v \sqcap [c, +\infty]$, and rename variables accordingly.
If the test uses two variables, e.g., $(v_1 < v_2)?$, then we create
intersections bound to {\em futures}.
We insert, at $l_f$, $v_{1f} = v_1 \sqcap [\fun{ft}(v_2), +\infty]$,
and $v_{2f} = v_2 \sqcap [-\infty, \fun{ft}(v_1) - 1]$.
Similarly, at $l_v$ we insert
$v_{1v} = v_1 \sqcap [-\infty, \fun{ft}(v_2) - 1]$
and $v_{2f} = v_2 \sqcap [\fun{ft}(v_1), +\infty]$.
We use the notation $\fun{ft}(v)$ to denote the {\em future} bounds of a
variable.
As we will show in Section~\ref{sub:micro}, once the growth pattern of $v$ is
known, we can replace $\fun{ft}(v)$ by an actual value.
Once we are done placing copies, we insert $\phi$-functions into the
transformed program to convert it to SSA form.
This last step avoids that two different names given to the same original
variable be simultaneously alive at the program code.
Figure~\ref{fig:ex_standard_eSSA}(a) shows our running example changed into
standard e-SSA form.
The part (b) of this figure shows the solution that we get to this new
program.
The e-SSA form allows us to bind interval information directly to the live
ranges of variables; thus, allowing us to solve range analysis sparsely.
More traditional approaches, which we call {\em dense analyses}, bind
interval information to pairs formed by variables and program points.

@@

\paragraph{Extracting Constraints.}
Our implementation handles 18 different assembly instructions.
The constraints in Figure~\ref{fig:eval_function} show some meaningful
examples.
These instructions include, for instance, the multiplicative operators
\texttt{mul}, \texttt{div} and \texttt{modulus},
the bitwise operators \texttt{and}, \texttt{or}, \texttt{xor} and \texttt{neg},
the different types of shifts, the additive operators \texttt{add} and
\texttt{sub}, and the logical operators \texttt{andalso}, \texttt{orelse} and
\texttt{not}.
We also handle different types of type conversion, e.g., converting 8-bits
characters to 32-bits integers and vice-versa.
Our constraint system deals with the $\phi$-functions from the SSA-form
representation.
Finally, we handle conditionals via intersection constraints created
during the live range splitting of variables.
The widening analysis that we will introduce in Section~\ref{sub:micro}
require monotonic transfer functions.
Many assembly operations, such as modulus or division, do not afford us
this monotonicity.
However, these non-monotonic instructions have conservative approximations.
%TODO: fix this discussion.

@@

\paragraph{The Constraint Graph.}
The main data structure that we use to solve the range analysis problem is
a variation of Ferrante {\em et al.}'s {\em program dependence
graph}~\cite{Ferrante87}.
For each constraint variable $V$ we create a variable node $N_v$.
For each constraint $C$ we create a constraint node $N_c$.
We add an edge from $N_v$ to $N_c$ if the name $V$ is used in $C$.
We add an edge from $N_c$ to $N_v$ if the constraint $C$ defines the name
$V$.
Figure~\ref{fig:ex_graph} shows the dependence graph that we build for the
e-SSA form program given in Figure~\ref{fig:ex_standard_eSSA}(a).
If $V$ is used by $C$ as the input of a future, then the edge from
$N_v$ to $N_c$ represents what Ferrante {\em et al.} call a {\em control
dependence}~\cite[p.323]{Ferrante87}.
We use dashed lines to represent these edges.
All the other edges denote {\em data dependences}~\cite[p.322]{Ferrante87}.

@@

\paragraph{Solving Range Analysis on Strong Components}
Having built the dependence graph, we proceed to solve the range analysis
problem by finding all the strongly connected components (SCCs) of the
dependence graph, and reducing this graph to a directed acyclic graph.
We perform this last step by collapsing every SCC into a single node.
We then sort the resulting DAG topologically, and apply the analyses from
Section~\ref{sub:micro} on every SSC in topological order.
Once we solve the range analysis problem for a SCC, we propagate the
intervals that we found to the variable nodes at the {\em frontier} of this
SCC.
A variable node $N_v$ is said to be in the frontier of a strongly connected
component $S$ if:
(i) $N_v \notin S$; and
(ii) there exists a variable node $N_v' \in S$, and a constraint node $N_c$,
such that $N_v' \leftarrow N_c$, and $N_c \leftarrow N_v$.
This propagation gives us a very modular algorithm.
It ensures that when analyzing a strongly connected component $S$ any influence
that $S$ might suffer from nodes outside it has been already
taken into consideration.

@@

\begin{figure}[t!]
\begin{center}
\begin{tabular}{c@{\hspace{1.5cm}}c}
\begin{minipage}{2cm}
\includegraphics{images/growth_lattice}
\end{minipage}
&
\begin{minipage}{8cm}
\begin{small}
\begin{eqnarray*}
\begin{array}{c@{\hspace{0.5cm}}c}
\inferrule{I[Y] = [\bot, \bot]}{I[Y] \leftarrow e(Y)}
&
\inferrule{\lb{e(Y)} < \lb{I[Y]} \\ \ub{e(Y)} > \ub{I[Y]}}
{I[Y] \leftarrow [-\infty, +\infty ]}
\\
\\
\inferrule{\lb{e(Y)} < \lb{I[Y]}}
{I[Y] \leftarrow [-\infty, \ub{I[Y]}]}
&
\inferrule{\ub{e(Y)} > \ub{I[Y]}}
{I[Y] \leftarrow [\lb{I[Y]}, +\infty]}
\end{array}
\end{eqnarray*}
\end{small}
\end{minipage}
\end{tabular}
\end{center}
\caption{\label{fig:growth_analysis}
(Left) The lattice of the growth analysis.
(Right) Cousot and Cousot's widening operator. We evaluate the rules from top 
to bottom, and stop upon finding a pattern matching.}
\end{figure}

\noindent
\textbf{Fixing futures.}
The ranges found by the widening analysis tells us which variables have fixed
bounds, independent on the intersections in the constraint system.
Thus, we can use actual limits to replace intersections bounded by futures.
Figure~\ref{fig:fix_intersects} shows the rules to perform these substitutions.
In order to correctly replace a future $\fun{ft}(V)$ that limits a variable
$V'$, we need to have already applied the widening analysis onto $V$.
Had we considered only data dependence edges, then it would be possible
that $V$ be analyzed before $V'$.
However, because of control dependence edges, this case cannot happen.
The control dependence edges ensure that any topological ordering of the
constraint graph either places $N_v$ before $N_{v'}$, or places these nodes
in the same strongly connected component.
For instance, in Figure~\ref{fig:ex_graph}, variables $J_1$ and $I_t$ are in
the same SCC only because of the control dependence edges.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We have implemented our range analysis algorithm in LLVM 3.0, and have run
experiments on a Intel quad core CPU with a 2.40GHz clock, and 3.6GB of RAM.
Each core has a 4,096KB L1 cache.
We use Linux Ubuntu 10.04.4.
Our implementation of range analysis has 3,958 lines of
commented C++ code, and our e-SSA conversion module has 672 lines.
We have analyzed the 428 C programs in the LLVM test suite, plus the integer
benchmarks in SPEC CPU 2006.
Together, these programs contain 4.76 million assembly instructions.

\noindent
\textbf{Structural Properties.}
Table~\ref{tab:struct} shows data taken from the SPEC 2006 integer benchmark
suite for the inter-procedural version of our analysis.
Three observations provide evidence that our algorithm is linear in practice,
as we will show in Figure~\ref{fig:TimeCorr}.
First, most of the strongly connected components found in constraint graphs are
singletons.
For instance, 99.11\% of the SSCs in gcc (\texttt{403.gcc}) have only
one node.
Hence, the most complex phases of our algorithm are confined to a small fraction
of the constraint graphs.
Second, the strongly connected components are not large -- our largest
component, found in \texttt{403.gcc}, has 2,131 nodes.
It exists due to a long chain of mutually recursive function calls.
We do not count control dependency edges when measuring the size of SCCs, as
they have no impact on the algorithm explained in Section~\ref{sub:micro}.
Third, as we see in the table, any node was visited at most three times by
the narrowing operator.

@@

\noindent
\textbf{The impact of whole program analysis on precision.}
Figure~\ref{fig:essaDivSSA}(Top) shows that the whole program analysis increases
moderately the precision of our analysis.
We show results for the five largest programs in three different categories of
benchmarks: SPEC CPU 2006, the Stanford Suite~\footnote{\texttt{http://classes.engineering.wustl.edu/cse465/docs/BCCExamples/stanford.c}} and
Bitwise~\cite{Stephenson00}.
Our initial goal when developing this analysis was to support a bitwidth-aware
register allocator.
Thus, we measure precision by the average number of bits that our
analysis allows us to save per program variables.
Our results for the SPEC programs were disappointingly: on the average, the
intra-procedural version of our analysis saves 5.23\% of the total program
bitwidth.
By using the inter-procedural version, with function inlining to simulate
context-sensitiveness, we can increase this number to 8.89\%.
A manual inspection of the SPEC programs reveal that this result is expected:
they manipulate files, and their code do no provide enough
explicit constants to power our analysis up.
However, with numerical benchmarks we fare much better.
On the average our inter procedural algorithm reduces the bitwidth of the
Stanford benchmarks by 36.24\%.
Notice that this number is better than the one we get with inlining (31.97\%),
because functions in those benchmarks are called only at one location.
Finally, for Bitwise we obtain a bitwidth reduction of 12.27\%.
However, this average is brought down by two outliers: \texttt{edge\_detect} and
\texttt{sha}, which cannot be reduced.
The bitwise benchmarks were implemented by Stephenson
{\em et al.}~\cite{Stephenson00} to validate their bitwidth analysis.
Our results are on par with those found by the original authors.
The bitwise programs contain only the \texttt{main} function; thus, different
versions of our algorithm find the same results.

>>>>

% Our contribution.
In this paper we provide a complete description of a range analysis algorithm,
and show extensive experimental data that justifies our engineering choices.
Our first algorithmic contribution on top of previous works is a three-phases
approach to handle comparisons between variables without resorting to any
exponential time technique.
The few publicly available implementations of range analyses that we are
aware of, such as those in FLEX~\footnote{The MIT's FLEX/Harpoon compiler
provides an implementation of Stephenson's algorithm~\cite{Stephenson00}, and
is available at \texttt{http://flex.cscott.net/Harpoon/}.} and
gcc~\footnote{Gcc's VRP pass implements a variant of Patterson's
algorithm~\cite{Patterson95}.} only deal with comparisons between variables and
constants.
Even theoretical works, such as Su and Wagner's~\cite{Su05} or Gawlitza
{\em et al.}'s~\cite{Gawlitza09} suffer from this limitation.
This deficiency is one of the reasons explaining why none of these works has
made their way into industrial-strength compilers.
Two other insights allow our implementation to scale up to very large
programs.
We use Bodik's Extended Static Single Assignment (e-SSA) form~\cite{Bodik00} to
perform range analysis sparsely.
This program representation ensures that the interval associated with
a variable is constant along its entire live range, and allows us to compute
path-sensitive results.
Finally, we process the strongly connected components that underline our
constraint system in topological ordering.
It is well-known that this technique is essential to speedup constraint solving
algorithms~\cite[Sec 6.3]{Nielson99}; however, we show that a careful
propagation of information along strong components not only gives us speed, but
also improves the precision of our results.

@@

In this section we explain the algorithm that we use to solve the range
analysis problem.
This algorithm involves a number of steps.
First, we convert the program to a suitable intermediate representation that
makes it easier to extract constraints.
From these constraints, we build a dependence graph that allows us to do
range analysis sparsely.
Finally, we solve the constraints applying different fix-point iterators on
this dependence graph.
Figure~\ref{fig:algorithm} gives a global view of this algorithm.
Some of the steps in the algorithm are optional.
They improve the precision of the range analysis, at the expense of a longer
running time.

@@

\noindent
\textbf{Choosing a Program Representation.}
The solution to the range analysis problem in Figure~\ref{fig:ex1}
is imprecise because we did not take conditional tests into considerations.
Branches give us information about the ranges that some variables assume, but
only at {\em specific} program points.
For instance, given a test such as $(k_1 < 100)?$ in  Figure~\ref{fig:ex1}(b),
we know that $I[k_1] \sqsubseteq [-\infty, 99]$ whenever the condition is true.
In order to encode this information, we might split the live range of $k_1$
right after the branching point; thus, creating two new variables, one at the
path where the condition is true, and another where it is false.
There is a program representation, introduced by Bodik
{\em et al.}~\cite{Bodik00}, that performs this live range splitting:
the {\em Extended Static Single Assignment} form, or e-SSA for short.

Given that the exact rules to convert a program to e-SSA form have never been
explicitly stated in the literature, we describe our rules as follows.
Let $(v < c)?$ be a conditional test, and let $l_t$ and $l_f$ be labels in
the program, such that $l_t$ is the target of the test if the condition is true,
and $l_f$ is the target when the condition is false.
We split the live range of $v$ at any of these points if at least one of two
conditions is true:
(i) $l_f$ or $l_t$ dominate any use of $v$;
(ii) there exist a use of $v$ at the dominance frontier of $l_f$ or $l_t$.
For the notions of dominance and dominance-frontier, see Aho
{\em et al.}~\cite{Aho06}.
To split the live range of $v$ at $l_f$ we insert at this
program point a copy $v_f = v \sqcap [c, +\infty]$, where $v_f$ is a fresh name.
We then rename every use of $v$ that is dominated by $l_f$ to $v_f$.
Dually, if we must split at $l_t$, then we create at this point a copy
$v_t = v \sqcap [-\infty, c-1]$, and rename variables accordingly.
If the conditional uses two variables, e.g., $(v_1 < v_2)?$, then we create
intersections bound to {\em futures}.
We insert, at $l_f$, $v_{1f} = v_1 \sqcap [\fun{ft}(v_2), +\infty]$,
and $v_{2f} = v_2 \sqcap [-\infty, \fun{ft}(v_1)]$.
Similarly, at $l_t$ we insert
$v_{1v} = v_1 \sqcap [-\infty, \fun{ft}(v_2) - 1]$
and $v_{2f} = v_2 \sqcap [\fun{ft}(v_1) + 1, +\infty]$.

We use the notation $\fun{ft}(v)$ to denote the {\em future} bounds of a
variable.
As we will show in Section~\ref{sub:micro}, once the growth pattern of $v$ is
known, we can replace $\fun{ft}(v)$ by an actual value.
Once we are done placing copies, we insert $\phi$-functions into the
transformed program to convert it to SSA form.
This last step avoids that two different names given to the same original
variable be simultaneously alive at the program code.
Figure~\ref{fig:ex_standard_eSSA}(a) shows our running example changed into
standard e-SSA form.
The part (b) of this figure shows the solution that we get to this new
program.
The e-SSA form allows us to bind interval information directly to the live
ranges of variables; thus, giving us the opportunity to solve range analysis
sparsely.
More traditional approaches, which we call {\em dense analyses}, bind
interval information to pairs formed by variables and program points.

% TODO: check with Victor ** eSSA description could also say that we create sigmas even if lt/lf does not dominate any use of v (under some conditions), i.e. when there is a use in dominance frontier inside a phi function **

@@

\noindent
\textbf{Extracting Constraints.}
Our implementation handles 18 different assembly instructions.
The constraints in Figure~\ref{fig:eval_function} show only a few examples.
Instructions that we did not show include, for instance, the multiplicative
operators \texttt{div} and \texttt{modulus},
the bitwise operators \texttt{and}, \texttt{or}, \texttt{xor} and \texttt{neg},
the different types of shifts, and the
logical operators \texttt{andalso}, \texttt{orelse} and \texttt{not}.
Most of these instructions are sign-agnostic; that is, given that numbers are
internally represented in 2's complement, the same implementation of a
constraint handles positive and negative numbers.
However, there are instructions that require different constraints, depending on
the input being signed or not.
Examples include \texttt{modulus} and \texttt{div}.
We also handle different kinds of type conversion, e.g., converting 8-bit
characters to 32-bit integers and vice-versa.
In addition to constraints that represent actual assembly instructions, we have
constraints to represent $\phi$-functions, and intersections, as seen in
Figure~\ref{fig:eval_function}.
The widening analysis that we will introduce in Section~\ref{sub:micro}
require monotonic transfer functions.
Many assembly operations, such as modulus or division, do not afford us
this monotonicity.
However, these non-monotonic instructions have conservative
approximations~\cite{Warren02}.

@@

\noindent
\textbf{The Constraint Graph.}
The main data structure that we use to solve the range analysis problem is
a variation of Ferrante {\em et al.}'s {\em program dependence
graph}~\cite{Ferrante87}.
For each constraint variable $V$ we create a variable node $N_v$.
For each constraint $C$ we create a constraint node $N_c$.
We add an edge from $N_v$ to $N_c$ if the name $V$ is used in $C$.
We add an edge from $N_c$ to $N_v$ if the constraint $C$ defines the name
$V$.
Figure~\ref{fig:ex_graph} shows the dependence graph that we build for the
e-SSA form program given in Figure~\ref{fig:ex_standard_eSSA}(a).
If $V$ is used by $C$ as the input of a future, then the edge from
$N_v$ to $N_c$ represents what Ferrante {\em et al.} call a {\em control
dependence}~\cite[p.323]{Ferrante87}.
We use dashed lines to represent these edges.
All the other edges denote {\em data dependences}~\cite[p.322]{Ferrante87}.

@@

\noindent
\textbf{Solving Range Analysis on Strong Components.}
We solve the range analysis
problem by finding all the strongly connected components (SCCs) of the
dependence graph, and reducing this graph to a directed acyclic graph.
We perform this last step by collapsing every SCC into a single node.
We then sort the resulting DAG topologically, and apply the analyses from
Section~\ref{sub:micro} on every SCC in topological order.
Once we solve the range analysis problem for a SCC, we propagate the
intervals that we found to the variable nodes at the {\em frontier} of this
SCC.
A variable node $N_v$ is said to be in the frontier of a strongly connected
component $S$ if:
(i) $N_v \notin S$; and
(ii) there exists a variable node $N_v' \in S$, and a constraint node $N_c$,
such that $N_v' \leftarrow N_c$, and $N_c \leftarrow N_v$.
This propagation gives us a very modular algorithm.
It ensures that when analyzing a strongly connected component $S$ any influence
that $S$ might suffer from nodes outside it has been already
taken into consideration.

@@

\begin{figure}[t!]
\begin{center}
\begin{tabular}{c@{\hspace{1.5cm}}c}
\begin{minipage}{2cm}
\includegraphics{images/growth_lattice}
\end{minipage}
&
\begin{minipage}{8cm}
\begin{small}
\begin{eqnarray*}
\begin{array}{c@{\hspace{0.5cm}}c}
\inferrule{I[Y] = [\bot, \bot]}{I[Y] \leftarrow e(Y)}
&
\inferrule{\lb{e(Y)} < \lb{I[Y]} \\ \ub{e(Y)} > \ub{I[Y]}}
{I[Y] \leftarrow [-\infty, +\infty ]}
\\
\\
\inferrule{\lb{e(Y)} < \lb{I[Y]}}
{I[Y] \leftarrow [-\infty, \ub{I[Y]}]}
&
\inferrule{\ub{e(Y)} > \ub{I[Y]}}
{I[Y] \leftarrow [\lb{I[Y]}, +\infty]}
\end{array}
\end{eqnarray*}
\end{small}
\end{minipage}
\end{tabular}
\end{center}
\caption{\label{fig:growth_analysis}
(Left) The lattice of the growth analysis.
(Right) Cousot and Cousot's widening operator. We evaluate the rules from
left-to-right, top-to-bottom, and stop upon finding a pattern matching.}
\end{figure}

\noindent
\textbf{Fixing futures.}
The ranges found by the widening analysis tells us which variables have fixed
bounds, independent on the intersections in the constraint system.
Thus, we can use actual limits to replace intersections bounded by futures.
Figure~\ref{fig:fix_intersects} shows the rules to perform these substitutions.
In order to correctly replace a future $\fun{ft}(V)$ that limits a variable
$V'$, we need to have already applied the widening analysis onto $V$.
Had we considered only data dependence edges, then it would be possible
that $V'$ be analyzed before $V"$.
However, because of control dependence edges, this case cannot happen.
The control dependence edges ensure that any topological ordering of the
constraint graph either places $N_v$ before $N_{v'}$, or places these nodes
in the same strongly connected component.
For instance, in Figure~\ref{fig:ex_graph}, variables $J_1$ and $I_t$ are in
the same SCC only because of the control dependence edges.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We have implemented our range analysis algorithm in LLVM 3.0, and have run
experiments on a Intel quad core CPU with a 2.40GHz clock, and 3.6GB of RAM.
Each core has a 4,096KB L1 cache.
We use Linux Ubuntu 10.04.4.
Our implementation of range analysis has 3,958 lines of
commented C++ code, and our e-SSA conversion module has 672 lines.
We have analyzed 428 C programs that constitute the LLVM test suite plus the
integer benchmarks in SPEC CPU 2006.
Together, these programs contain 4.76 million assembly instructions.

\noindent
\textbf{Structural Properties.}
Table~\ref{tab:struct} shows data taken from the SPEC 2006 integer benchmark
suite for the inter-procedural version of our analysis.
Three observations provide evidence that our algorithm is linear in practice,
as we will show in Figure~\ref{fig:TimeCorr}.
First, most of the strongly connected components found in constraint graphs are
singletons.
For instance, 99.11\% of the SCCs in gcc (\texttt{403.gcc}) have only
one node.
Hence, the most complex phases of our algorithm are confined to a small part
of the constraint graph.
Second, the strongly connected components are not large -- our largest
component, found in \texttt{403.gcc}, has 2,131 nodes.
It exists due to a long chain of mutually recursive function calls.
We do not count control dependency edges when measuring the size of SCCs, as
they have no impact on the algorithm explained in Section~\ref{sub:micro}.
Third, as we see in the table, any node was visited at most three times by
the narrowing operator.

@@

\noindent
\textbf{The impact of whole program analysis on precision.}
Figure~\ref{fig:essaDivSSA}(Top) shows that the whole program analysis increases
moderately the precision of our analysis.
We show results for the five largest programs in three different categories of
benchmarks: SPEC CPU 2006, the Stanford Suite~\footnote{\texttt{http://classes.engineering.wustl.edu/cse465/docs/BCCExamples/stanford.c}} and
Bitwise~\cite{Stephenson00}.
Our initial goal when developing this analysis was to support a bitwidth-aware
register allocator.
Thus, we measure precision by the average number of bits that our
analysis allows us to save per program variables.
It is very important to notice that we do not consider the constants in the
program text in our statics of precision.
In other words, we only measure bitwidth reduction in variables that a constant
propagation step could not remove.
Our results for the SPEC programs were disappointingly: on the average, the
intra-procedural version of our analysis saves 5.23\% of the total program
bitwidth.
By using the inter-procedural version, with function inlining to simulate
context-sensitiveness, we can increase this number to 8.89\%.
A manual inspection of the SPEC programs reveal that this result is expected:
they manipulate files, and their code do no provide enough
explicit constants to power our analysis up.
However, with numerical benchmarks we fare much better.
On the average our inter-procedural algorithm reduces the bitwidth of the
Stanford benchmarks by 36.24\%.
Notice that this number is better than the one we get with inlining (31.97\%),
because functions in those benchmarks are called only at one location.
Finally, for Bitwise we obtain a bitwidth reduction of 12.27\%.
However, this average is brought down by two outliers: \texttt{edge\_detect} and
\texttt{sha}, which cannot be reduced.
The bitwise benchmarks were implemented by Stephenson
{\em et al.}~\cite{Stephenson00} to validate their bitwidth analysis.
Our results are on par with those found by the original authors.
The bitwise programs contain only the \texttt{main} function; thus, different
versions of our algorithm find the same results.
